---
title: "p8105_hw3_fl2715"
author: "Fengwei Lei"
output: github_document
---

## Loading Library and Settings
```{r}
library(tidyverse)
library(ggridges)
library(patchwork)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 9, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Problem 1

### Loading Dataset
```{r}
library(p8105.datasets)
data("ny_noaa")
```

### A Brief Data Discription

```{r}
# Summarize missing values
ny_noaa  |> 
  summarize(across(everything(), ~sum(is.na(.))))  |> 
  pivot_longer(cols = everything(), names_to = "variable", values_to = "missing_values")  |> 
  arrange(desc(missing_values))
```

The dataset contains daily weather observations from various NOAA stations in New York, with `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns. The key variables include date (date of observation), prcp (precipitation (tenths of mm)), snow (Snowfall (mm)), snwd (snow depth (mm)), tmax (maximum temperature (tenths of degrees C)) and tmin (Minimum temperature (tenths of degrees C)). Missing data is present in the above table, particularly temperature and precipitation measurements. So we need to deal with missing values issue first for further analysis.

### Answering Given Questions

#### Data Cleaning

We do some data cleaning process. It begins by converting the date variable into individual year, month, and day variables using the `separate()` function. Following this, we convert `tmax` and `tmin` to numeric.
```{r}
ny_noaa_clean = ny_noaa  |> 
  mutate(date = as.character(date)) |> 
  separate(date, into = c("year", "month", "day"), convert=TRUE)  |> 
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin),
    tmax= tmax /10,
    tmin=tmin/10)
```

After that, we find the most commonly observed values for snowfall.
```{r}
ny_noaa_clean  |>
  count(snow)  |> 
  arrange(desc(n))  |> 
  head(10)
```

The most commonly observed snowfall value is 0, with 2,008,508 occurrences. This high frequency of zero snowfall indicates that on many days, no snowfall was recorded. This is common because snowfall occurs only during specific weather conditions in NY. The second most commonly observed value is `NA`, showing missing values. 

#### Plot for Average Maximum Temperature in January and July

The code creates a two-panel plot showing the average maximum temperature in January and July across multiple weather stations over the years. It groups the data by station ID, year, and month, then calculates the mean maximum temperature (mean_tmax) for January and July, using `facet_grid(~month)` to display separate panels for each month. 
The plot reveals expected seasonal trends: January’s average temperatures are consistently lower than July’s for all stations and years. The stations tend to show similar temperature fluctuations across years, with peaks and valleys aligned. Notably, one station in July 1987 or 1988 shows an unusually low temperature, along with a few other minor outliers.

```{r}
ny_noaa_clean  |> 
  group_by(id, year, month)  |> 
  filter(month %in% c(1, 7), !is.na(tmax))  |> 
  summarize(mean_tmax = mean(tmax, na.rm = TRUE, color = id)) |> 
  ggplot(aes(x = year, y = mean_tmax, group = id)) + geom_point() + geom_path() +
  facet_grid(~month) +
  labs(
    title = "Average Max Temperature in January and July Across Stations",
    x = "Year", y = "Average Max Temperature"
  )
```

#### Two Plots for "tmax vs tmin" and "Snowfall distribution for values > 0 and < 100"

```{r}
p1 = ny_noaa_clean  |> 
  filter(!is.na(tmax), !is.na(tmin))  |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex()
  labs(
    title = "tmax vs tmin Plot",
    x = "Minimum Temperature (°C)", y = "Maximum Temperature (°C)"
  )


p2 = ny_noaa_clean  |> 
  filter(snow > 0 & snow < 100)  |> 
  ggplot(aes(x = snow, y = factor(year))) +
  geom_density_ridges(scale = .85)+
  labs(
    title = "Distribution of Snowfall (0 < snow < 100 mm) by Year",
    x = "Year", y = "Snowfall (mm)"
  ) 

p1 + p2
```

## Problem 2

### Load and Tidy the Data

Firstly, we load the original demographic and accelerometer data separately. After that, we merge these two dataset, following with some tidy process. It includes excluding participants less than 21 years of age and missing demographic data, pivoting the data form to longer form and encoding data with reasonable variable classes.

```{r}
## Load the demographic and accelerometer data
demo_data=read.csv(
  "data/nhanes_covar.csv",skip=4) |> 
  janitor::clean_names()

acceler_data=read.csv(
  "data/nhanes_accel.csv") |> 
  janitor::clean_names()

#merge and tidy the data
data=merge(
  demo_data, acceler_data, by="seqn") |> 
  drop_na() |> 
  filter(age >=21) |> 
  pivot_longer(
    cols=min1:min1440,
    names_to = "time",
    values_to = "mims",
    names_prefix ="min") |> 
 mutate(
    sex = factor(sex, levels = c(1, 2), labels = c("male", "female")),
    education = factor(education, levels = c(1, 2, 3), labels = c("less than high school", "high school equivalent", "more than high school"))
  )
```

### Table: Number of Men and Women in Each Education Category

We produce a reader-friendly table for the number of men and women in each education category.

```{r}
education_sex_table = data  |> 
  group_by(education, sex)  |> 
  summarize(count = n_distinct(seqn))  |> 
  pivot_wider(names_from = sex, values_from = count, values_fill = 0)

knitr::kable(education_sex_table, caption = "Number of Men and Women in Each Education Category")
```
Comments:

The above table presents in the “less than high school” group, there are slightly more women (28) than men (27), indicating a near balance in this educational attainment level. In the “high school equivalent” category, the number of men (35) exceeds that of women (23), highlighting a gender disparity in this education level. Conversely, in the “more than high school” category, the numbers for men (56) and women (59) are closely matched, with women slightly outnumbering men. The data suggests that while there are some gender imbalances in educational attainment, particularly at the high school level, the distribution becomes more equitable at higher education levels.

### Age Distribution Visualization

We create a visualization of the age distributions for men and women in each education category. 

```{r}
data_unique = data  |> 
  distinct(seqn, .keep_all = TRUE)

ggplot(data_unique, aes(x = age, fill = sex)) +
  geom_density(alpha = 0.5) +
  facet_grid(~ education) +
  labs(title = "Age Distribution by Education and Sex", x = "Age", y = "Density")
```

Comments:

In all three panels, females (yellow) generally outnumber males (purple) at older ages for those with higher education or less. Also, for those with high school equivalent or less education level, the age distribution of males peaks earlier than females. In the "more than high school" group, the distributions of male and female both peak at about age 31, while male dominate at older ages (60+).

### Total Activity vs. Age Plot

We aggregate across minutes to create a total activity variable for each participant. And we plot these total activities (y-axis) against age (x-axis), with comparing men to women and having separate panels for each education level. Also, we include a trend line or a smooth to illustrate differences.

```{r}
total_activity = data  |> 
  group_by(seqn, sex, education, age)  |> 
  summarise(total_mims = sum(mims, na.rm = TRUE))

# Plot total activity vs. age, comparing men and women, and faceting by education level
ggplot(total_activity, aes(x = age, y = total_mims, color = sex)) +
  geom_point() +
  geom_smooth(method = "lm",se = FALSE) +
  facet_wrap(~ education) +
  labs(title = "Total Activity vs. Age by Education and Sex", x = "Age", y = "Total Activity (MIMS)") +
  theme_minimal()
```

Comments:

In all categories, physical activity declines with age, but the decline is steeper for female (yellow) especially for those with less than a high school education. Females (yellow) seem to maintain higher activity levels later in life, particularly among those with more education. There is greater variability in total activity among older individuals in all groups, with distinct linear trends highlighting the sex differences.

### 24-Hour Activity Time Course Plot

Now we make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex.

```{r}
hourly_data = data  |> 
  mutate(hour = as.numeric(time) / 60) |> 
  group_by(education, sex, hour)  |> 
  summarize(mean_mims = mean(mims), .groups = "drop")

ggplot(hourly_data, aes(x = hour, y = mean_mims, color = sex)) +
  geom_point(alpha = .2) + 
  geom_smooth(se = FALSE, aes(group = sex)) +
  facet_wrap(~ education, ncol = 1) +
  labs(title = "24-Hour Activity Time Courses by Education Level and Sex",
       x = "Hour of the Day",
       y = "Mean MIMS",
       color = "Sex")
```

Conclusion:

This graph presents the average physical activity across a 24-hour period, divided by education level and sex. Across all education levels, activity peaks about between 7:00 and 18:00, with males (purple) and females (yellow) showing similar patterns. However, females consistently demonstrate higher activity levels throughout the day, especially in the group with more than a high school education. As expected, the night-time hours show very low activity. The overall trends suggest that daily activity follows similar patterns regardless of education level, but females generally exhibit higher levels of physical activity across the board.

## Problem 3

### Import, Clean, and Tidy Data
```{r}
#Import data
jan_2020_data=read_csv(
  "data/Jan 2020 Citi.csv") |> 
  janitor::clean_names() |> 
  mutate(
    year=2020,
    month="January")

july_2020_data=read_csv(
  "data/July 2020 Citi.csv") |> 
  janitor::clean_names() |> 
  mutate(
    year=2020,
    month="July")

jan_2024_data=read_csv(
  "data/Jan 2024 Citi.csv") |> 
  janitor::clean_names() |> 
  mutate(
    year=2024,
    month="January")

july_2024_data=read_csv(
  "data/July 2024 Citi.csv") |> 
  janitor::clean_names() |> 
  mutate(
    year=2024,
    month="July")

# merge the data
combined_data = bind_rows(jan_2020_data, july_2020_data, jan_2024_data, july_2024_data)
```

